---
  title: "Getting Started with BayesBadger"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BayesBadger}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---
  
  ```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # set to TRUE if you want code to run on build
)
```

## Overview

**BayesBadger** fits a Bayesian double generalized beta regression model with
spatial random effects. It is designed for outcomes that are proportions or
rates strictly between 0 and 1, where both the mean and the dispersion of the
response may depend on covariates, and where observations are grouped into
spatial units with a known adjacency structure.

The model has three components:
  
  - **Mean model**: individual-level covariates (binary, continuous, categorical)
linked to the mean response via a logit link.
- **Dispersion model**: spatial-unit-level covariates linked to the dispersion
parameter via a log link.
- **Spatial random effects**: a graph Laplacian prior encodes spatial smoothness,
so nearby units share information about their random effects.

Sampling is performed via Stan using `rstan`.

---
  
  ## Installation
  
  ```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("yourusername/BayesBadger")
```

---
  
  ## Setup
  
  ```{r libraries}
library(BayesBadger)
library(rstan)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

set.seed(42)
```

---
  
  ## Step 1: Define the Spatial Structure
  
  We simulate 100 spatial units arranged on a 10×10 grid. Each unit is connected
to its immediate horizontal and vertical neighbours, forming a regular lattice.
The adjacency matrix must be symmetric with named rows and columns.

```{r spatial-structure}
S        <- 100
grid_dim <- 10
unit_names <- paste0("UNIT_", sprintf("%03d", 1:S))

A_mat <- matrix(0, S, S, dimnames = list(unit_names, unit_names))
node_id <- function(r, c) (r - 1) * grid_dim + c

for (r in 1:grid_dim) {
  for (c in 1:grid_dim) {
    s <- node_id(r, c)
    if (c < grid_dim) {
      nb <- node_id(r, c + 1)
      A_mat[s, nb] <- A_mat[nb, s] <- 1
    }
    if (r < grid_dim) {
      nb <- node_id(r + 1, c)
      A_mat[s, nb] <- A_mat[nb, s] <- 1
    }
  }
}

cat(sprintf("Adjacency matrix: %d x %d,  mean degree = %.2f\n",
            S, S, mean(rowSums(A_mat))))
```

---
  
  ## Step 2: Set True Parameters
  
  We define the true parameter values used to simulate data. These include
coefficients for the mean model (`beta`), the dispersion model (`omega`),
and spatially structured random effects (`alpha`).

```{r true-parameters}
# Mean model coefficients
beta_true <- c(
  female   =  0.25,
  score    =  0.30,
  educmid  =  0.15,
  educhigh =  0.40
)

# Dispersion model coefficients
omega_true <- c(
  intercept =  1.20,
  density   =  0.30,
  income    = -0.20
)

# Spatial random effects: smooth 2-D wave rescaled to [-0.5, 0.5]
row_id <- ((1:S) - 1) %/% grid_dim + 1
col_id <- ((1:S) - 1) %%  grid_dim + 1

alpha_raw  <- sin(2 * pi * row_id / grid_dim) * cos(2 * pi * col_id / grid_dim)
alpha_true <- -0.5 + (alpha_raw - min(alpha_raw)) /
  (max(alpha_raw) - min(alpha_raw))

cat(sprintf("alpha_true range: [%.3f, %.3f]\n", min(alpha_true), max(alpha_true)))
```

---
  
  ## Step 3: Simulate Spatial-Unit-Level Data
  
  Each spatial unit has two dispersion covariates: population density and median
income. These are standardised and used to compute the true dispersion parameter
`phi` for each unit.

```{r spatial-data}
density_s <- as.numeric(scale(runif(S, 100, 5000)))
income_s  <- as.numeric(scale(rnorm(S, 50000, 15000)))

spatial_df <- data.frame(
  unit    = unit_names,
  density = density_s,
  income  = income_s
)

# True log-dispersion and phi per spatial unit
log_phi_true <- omega_true["intercept"] +
  omega_true["density"] * density_s +
  omega_true["income"]  * income_s
phi_true <- exp(log_phi_true)

cat(sprintf("Spatial-unit phi range: [%.2f, %.2f]\n",
            min(phi_true), max(phi_true)))
```

---
  
  ## Step 4: Simulate Individual-Level Data
  
  Individuals are nested within spatial units. Unit sizes are drawn from a
Poisson(5) distribution, with roughly 4% of units forced to be empty.
Empty units still appear in the adjacency matrix and spatial data — they
receive spatial random effect draws informed by their neighbours.

```{r individual-data}
# Unit sizes
n_per_unit <- rpois(S, lambda = 5)
n_per_unit <- pmax(n_per_unit, 1)

# Force ~4% of units to be empty
empty_units <- sample(seq_len(S), size = round(0.04 * S), replace = FALSE)
n_per_unit[empty_units] <- 0L

cat(sprintf("Total individuals: %d\n", sum(n_per_unit)))
cat(sprintf("Empty units: %d (%.1f%%)\n",
            sum(n_per_unit == 0), 100 * mean(n_per_unit == 0)))

# Repeat unit label for each individual
unit_id <- rep(unit_names, times = n_per_unit)
n       <- length(unit_id)

# Individual-level covariates
female <- rbinom(n, 1, 0.52)
score  <- as.numeric(scale(rnorm(n, 50, 10)))
educ   <- factor(
  sample(c("low", "mid", "high"), n, replace = TRUE, prob = c(0.35, 0.40, 0.25)),
  levels = c("low", "mid", "high")
)

# Simulate Beta response
X_sim   <- model.matrix(~ female + score + educ)[, -1, drop = FALSE]
alpha_i <- alpha_true[match(unit_id, unit_names)]
eta_mu  <- X_sim %*% beta_true + alpha_i
mu_i    <- 1 / (1 + exp(-eta_mu))
phi_i   <- phi_true[match(unit_id, unit_names)]

y_sim <- rbeta(n, shape1 = mu_i * phi_i, shape2 = (1 - mu_i) * phi_i)
y_sim <- pmax(pmin(y_sim, 1 - 1e-6), 1e-6)

indiv_df <- data.frame(
  unit   = unit_id,
  y      = y_sim,
  female = female,
  score  = score,
  educ   = educ
)

summary(indiv_df)
```

---
  
  ## Step 5: Fit the Model
  
  The formula uses `|` to separate the mean model (left) from the dispersion
model (right). Variables on the left are individual-level; variables on the
right are spatial-unit-level and must appear in `spatial_data`.

> **Note:** This model inverts a 100×100 matrix at each iteration. Allow
> approximately 10–20 minutes for fitting with the default settings.

```{r fit-model}
fit <- bayes_badger(
  formula          = y ~ female + score + factor(educ) | density + income,
  individual_data  = indiv_df,
  spatial_id       = "unit",
  spatial_data     = spatial_df,
  adjacency_matrix = A_mat,
  # Priors
  beta_prior_mean    = 0,
  beta_prior_var     = 10,
  omega_prior_mean   = 0,
  omega_prior_var    = 10,
  lambda_prior_shape = 10,
  lambda_prior_rate  = 1,
  gamma_prior_shape  = 1,
  gamma_prior_rate   = 10,
  # MCMC
  chains = 4,
  iter   = 2000,
  warmup = 1000,
  thin   = 1,
  seed   = 2024
)
```

The `print` method displays a summary of the posterior for the global
parameters. Spatial random effects (`alpha`) are omitted for brevity but
can be accessed via `rstan::summary(fit$stanfit)`.

```{r print-fit}
print(fit)
```

---
  
  ## Step 6: Marginal Effects for Mean-Model Covariates
  
  `marginal_effects()` computes the average marginal effect (AME) for each
mean-model predictor. For continuous variables the AME is the average change
in the predicted mean for a one-unit increase. For binary variables it is the
average change when the variable switches from 0 to 1.

The function returns a vector of posterior draws for each variable, which can
be summarised to obtain posterior means and credible intervals.

```{r marginal-effects-covariates}
me_covar <- marginal_effects(
  fit,
  variables = fit$mean_terms,
  type      = c(female   = "binary",
                score    = "continuous",
                educmid  = "binary",
                educhigh = "binary"),
  method = "mean",
  burnin = 0,
  thin   = 5
)

# Summarise posterior draws with HDI credible intervals
summarise_draws <- function(draw_list) {
  do.call(rbind, lapply(names(draw_list), function(nm) {
    x  <- draw_list[[nm]]
    ci <- HDInterval::hdi(x, credMass = 0.95)
    data.frame(
      variable = nm,
      mean     = round(mean(x), 4),
      lower_95 = round(ci["lower"], 4),
      upper_95 = round(ci["upper"], 4),
      row.names = NULL
    )
  }))
}

summarise_draws(me_covar)
```

---
  
  ## Step 7: Marginal Effects for Spatial Units
  
  `marginal_effects_spatial()` computes the AME for each spatial unit, defined
as the average change in the predicted mean when that unit's spatial random
effect replaces the cross-unit mean random effect. This quantifies how much
each unit's location shifts predicted outcomes relative to the spatial average.

The function returns an `n_draws × S` matrix. Each column corresponds to one
spatial unit and contains its posterior AME draws.

```{r marginal-effects-spatial}
me_spatial <- marginal_effects_spatial(
  fit,
  method = "mean",
  burnin = 0,
  thin   = 5
)

cat(sprintf("Output dimensions: %d draws x %d spatial units\n",
            nrow(me_spatial), ncol(me_spatial)))

# Summarise across draws for each spatial unit
me_spatial_summary <- data.frame(
  spatial_unit = colnames(me_spatial),
  mean_est     = round(colMeans(me_spatial), 4),
  sd           = round(apply(me_spatial, 2, sd), 4),
  `2.5%`       = round(apply(me_spatial, 2, quantile, probs = 0.025), 4),
  `97.5%`      = round(apply(me_spatial, 2, quantile, probs = 0.975), 4),
  check.names  = FALSE,
  row.names    = NULL
)

head(me_spatial_summary, 10)
```
