---
title: "Getting Started with BayesBadger"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BayesBadger}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE   # set to TRUE if you want code to run on build
)
```

## Overview

**BayesBadger** is an R package that implements a Bayesian double generalized
beta regression model with the mean modeled using individual-level covariates
and the precision (dispersion) modeled with cluster-level covariates. The
function `bayes_badger()` implements this model, and also allows for spatial
smoothing through a graph Laplacian prior on the cluster effects.
Additionally, the `bayes_badger()` function allows users to specify the mean
of the prior of the cluster effects, which can be used to incorporate temporal
borrowing into the model in settings with multiple time points. The package
also includes functions to compute marginal effects for both mean-model
covariates and clusters, which offer cleaner interpretation of the model's
results.

The model has three components:

- **Mean model**: individual-level covariates (binary, continuous, categorical)
  linked to the mean response via a logit link.
- **Precision model**: cluster-level covariates linked to the precision
  parameter via a log link.
- **Cluster effects**: cluster-level random effects, can be optionally estimated through a graph
  Laplacian prior that encodes spatial smoothness.

Sampling is performed via Stan using `rstan`.

---

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("abhijainstats/BayesBadger")
```

---

## Setup

```{r libraries}
library(BayesBadger)
library(rstan)
library(knitr)
library(kableExtra)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

---

## Example Data

`BayesBadger` comes with a simulated example dataset to illustrate the
modeling workflow. The data represent 100 clusters arranged on a 10×10 grid,
with individuals nested within clusters. The bundled objects are:

- **`indiv_df`** — individual-level observations with response `y` and
  covariates `female`, `inc`, and `educ`. 
- **`spatial_df`** — cluster-level covariates (`diversity`, `ses`) used for
  the dispersion model.
- **`A_mat`** — a 100×100 symmetric adjacency matrix encoding the
  neighbourhood structure of the 10×10 grid.
- The responde `y` is a continuous variable
  simulated from a double generalized beta distribution, with `female`, `inc`, and `educ` as mean-model covariates as well as a
  spatially structured cluster effect.

```{r load-data, echo=TRUE}
data(indiv_df)
data(spatial_df)
data(A_mat)

kable(head(indiv_df))
kable(head(spatial_df))
```

See `?indiv_df`, `?spatial_df`, and `?A_mat` for full documentation.

---

## Fit the Model

The formula uses `|` to separate the mean model (left) from the dispersion
model (right). Variables on the left are individual-level; variables on the
right are cluster-level and must appear in `cluster_data`.

```{r fit-model}
fit <- bayes_badger(
  formula          = y ~ female + inc + factor(educ) | diversity + ses,
  individual_data  = indiv_df,
  cluster_id       = "unit",
  cluster_data     = spatial_df,
  adjacency_matrix = A_mat,
  # Priors
  beta_prior_mean    = 0,
  beta_prior_var     = 10,
  omega_prior_mean   = 0,
  omega_prior_var    = 10,
  lambda_prior_shape = 10,
  lambda_prior_rate  = 1,
  gamma_prior_shape  = 1,
  gamma_prior_rate   = 10,
  # MCMC
  chains = 1,
  iter   = 2000,
  warmup = 1000,
  thin   = 1,
  seed   = 2024
)
```

You can inspect the fitted model by accessing its components, e.g. the
posterior summary for global parameters via `rstan::summary(fit$stanfit)`.

```{r inspect-fit, echo=TRUE, results='asis'}
kable(rstan::summary(fit$stanfit, pars = c("beta", "omega", "lambda", "gamma"))$summary)
```

---

## Marginal Effects for Mean-Model Covariates

`marginal_effects()` computes the average marginal effect (AME) for each
mean-model predictor. For continuous variables the AME is the average change
in the predicted mean for a one-unit increase. For binary variables it is the
average change when the variable switches from 0 to 1. The function returns
a vector of posterior draws for each variable, which can be summarized to
obtain posterior means and credible intervals.

```{r marginal-effects-covariates, echo=TRUE, results='asis'}
me_covar <- marginal_effects(
  fit,
  variables = fit$mean_terms,
  type      = c(female   = "binary",
                inc      = "continuous",
                educmid  = "binary",
                educhigh = "binary"),
  method = "mean",
  burnin = 0,
  thin   = 5
)

# Summarise posterior draws with HDI credible intervals
summarise_draws <- function(draw_list) {
  do.call(rbind, lapply(names(draw_list), function(nm) {
    x  <- draw_list[[nm]]
    ci <- HDInterval::hdi(x, credMass = 0.95)
    data.frame(
      variable = nm,
      mean     = round(mean(x), 4),
      lower_95 = round(ci["lower"], 4),
      upper_95 = round(ci["upper"], 4),
      row.names = NULL
    )
  }))
}

library(kableExtra)

knitr::kable(
  summarise_draws(me_covar),
  caption = "Posterior Marginal Effects for Mean-Model Covariates"
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

---

## Marginal Effects for Clusters

`marginal_effects_cluster()` computes the AME for each cluster, defined as
the average change in the predicted mean when that cluster's random effect
replaces the mean cluster effect. This quantifies how much belonging to a
particular cluster changes the predicted mean response, relative to the
average cluster effect across all clusters. The function returns an
`n_draws × S` matrix, where each column corresponds to one cluster and
contains its posterior AME draws.

```{r marginal-effects-cluster, echo=TRUE, results='asis'}
me_cluster <- marginal_effects_cluster(
  fit,
  method = "mean",
  burnin = 0,
  thin   = 5
)

# Summarise across draws for each cluster
me_cluster_summary <- data.frame(
  cluster     = colnames(me_cluster),
  mean_est    = round(colMeans(me_cluster), 4),
  sd          = round(apply(me_cluster, 2, sd), 4),
  `2.5%`      = round(apply(me_cluster, 2, quantile, probs = 0.025), 4),
  `97.5%`     = round(apply(me_cluster, 2, quantile, probs = 0.975), 4),
  check.names = FALSE,
  row.names   = NULL
)

kable(head(me_cluster_summary, 10))
```
